---
title: 'Spatial CODA'
author: "Supplemental material"
date: "Thi Huong An Nguyen, Anne Ruiz-Gazen, Christine Thomas-Agnan, Thibault Laurent"
output:
  html_document:
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
---
<link href="markdown7.css" rel="stylesheet">

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We provide the data and the **R** code used in the article "Spatial CODA" so that readers may reproduce all the figures, tables and statistics presented in the article with the **R** software. 

If you use this code, please cite: 

Nguyen T.H.A, Ruiz-Gazen, A., Thomas-Agnan C. and T. Laurent (2019). Spatial CODA. *WP*.

# Prerequisites #

Required packages:
```{r, eval = F}
install.packages(c("compositions", "mvnfast", "quantmod", "plot3D", "sp"))
```

Loading packages:
```{r, message = F}
require("classInt") # discretize numeric variable
require("compositions") # compositional data
require("ggplot2") # ggplot functions
require("mvnfast") # multivariate Student distribution 
require("quantmod")  # import financial data
require("plot3D") # plot distribution in 3D
require("RColorBrewer") # palette colors with R
require("rgdal") # import spatial data
require("sp") # spatial data
require("spdep") # spatial econometric modelling
```

Information about the current R session :
```{r}
sessionInfo()
```


# Simulation study #

This section demonstrates how to obtain the results presented in the section 3 of the article. We first present our  functions which can be adapted to another framework different from our simulation process.

## Simulation of spatial multivariate $Y$  ## 

The function *simu_spatial_multi_y()* simulates a multivariate $Y$ of the form $Y= Y\Gamma + WYR + X\beta + \epsilon$ where $\epsilon$ follows either a multivariate Gaussian (**method_simulate = "N"**), or the Independent multivariate Student (**method_simulate = "IT"**) distributions.  

Input arguments are :

* **X**, the matrix of explanatory variables of size $n \times K$,

* **beta_true**, the $\beta$ matrix of size $K\times L$ : 

$$\left(\begin{array}{ccc} 
\beta_{11} & \ldots & \beta_{1L} \\ 
\beta_{21} & \ldots & \beta_{2L} \\ 
\vdots & & \vdots \\ 
\beta_{K1} & \ldots & \beta_{KL}
\end{array}\right)$$

* **method_simulate**, the method of simulation (a character among "N", "IT"),

* **Sigma**, the matrix of size $L \times L$,

* **GAMMA**, the matrix of size $L \times L$,

* **RHO**, the matrix of size $L \times L$,

* **W**, the matrix of size $n \times n$,

* **nu**, for Student distribution only.

The function returns a matrix of size $n \times L$. To load the function:
  
```{r}
source("./R/simu_spatial_multi_y.R")
```


## Examples ##

### Preparation of the data ### 

Import the [Midi-Pyrénées communes boundaries](http://www.thibault.laurent.free.fr/code/GLT/contours.zip) into **R** which was used in Goulard et al. (2017): 
  
```{r,echo=FALSE,results="hide"}
URL <- "http://www.thibault.laurent.free.fr/code/GLT/"
fil <- "contours.zip"
if (!file.exists(fil)) download.file(paste(URL, fil, sep = ""), fil)
unzip(fil)
```

```{r,echo=TRUE,results="hide"}
mapMAP <- readOGR(dsn = "contours", layer = "ADTCAN_region")
```

We convert the type of the identification units into numeric values:
```{r}
mapMAP@data$CODE <- as.numeric(as.character(mapMAP@data$CODE))
```

The number of observations equals to $n$:
```{r}
n <- nrow(mapMAP)
```

We consider one spatial weight matrix $W$, based on the 10-nearest neighbours and row-normalized. $W$ is relatively sparse ($96.5\%$ of null values).
```{r,echo=TRUE}
coords <- coordinates(mapMAP)
W1.listw <- nb2listw(knn2nb(knearneigh(coords, 10)), 
                     style = "W")
W_simu <- listw2mat(W1.listw)
```

### Simulation of a multivariate SAR process ###

#### Example when $L = 2$ ####
We plan to simulate a multivariate $Y$ of size $L=2$ : 
  
```{r}
L_simu <- 2
```

We simulate the explanatory variables:
  
```{r}
set.seed(1234)
x1 <- rnorm(n, 15, 3)
x2 <- rbinom(n, 100, 0.45)/100
x3 <- log(round(runif(n, 1, n),0))
x_simu <- cbind(rep(1, n), x1, x2, x3)
p_simu <- ncol(x_simu)
```


The $\beta$ matrix is $$\left(\begin{array}{cc} 5 & 2\\ 1/4 & -1 \\ 6 & -3\\ 1 & 3\end{array}\right),$$ 
  
```{r}
beta_true <- matrix(c(5, 2, 0.25, -1, 6, -3, 1, 3), byrow = T,
                    nrow = p_simu, ncol = L_simu)
```

the $\Sigma$ matrix is  $$\left(\begin{array}{cc} 10 & 8\\ 8 & 10\end{array}\right),$$ 
  
```{r}
Sigma <- matrix(c(10, 9, 9, 10),
                nrow = L_simu, ncol = L_simu)
```

the $\Gamma$ matrix is $$\left(\begin{array}{cc} 0 & 0\\ 0 & 0\end{array}\right),$$ 
  
```{r}
GAMMA <- matrix(c(0, 0, 0, 0),
                nrow = L_simu, ncol = L_simu)
```

and the $R$ matrix is $$\left(\begin{array}{cc} 0.75 & 0.1\\ 0.3 & 0.15\end{array}\right).$$ 
  
```{r}
RHO <- matrix(c(0.5, 0.1, 0.2, 0.4),
              nrow = L_simu, ncol = L_simu)
```


We simulate the process:
  
```{r}
set.seed(1)
y_N <- simu_spatial_multi_y(X = x_simu, beta_true = beta_true, method_simulate = "N",
                            Sigma = Sigma, GAMMA = GAMMA, RHO = RHO, W = W_simu) 
mapMAP@data[, c("y_1", "y_2")] <- y_N
```

We plot the two component of $Y$ on the map: 
```{r, fig.width = 12, fig.height = 6}
library("cartography")
op <- par(mfrow = c(1, 2), oma = c(0, 0, 0, 0), mar = c(0, 0, 1, 0))
choroLayer(spdf = mapMAP, var = "y_1", legend.pos = "topleft", 
           method = "quantile")
choroLayer(spdf = mapMAP, var = "y_2", legend.pos = "topleft", 
           method = "quantile")
par(op)
```

We also plot the joint distribution:
```{r}
y_N_df <- data.frame(y_1 = y_N[, 1], y_2 = y_N[, 2])
sp <- ggplot(y_N_df, aes(x = y_1, y = y_2)) +
  geom_point()
sp + geom_density_2d()
```

#### Example when $L = 3$ ####

We simulate another multivariate sample when $L = 3$.

```{r}
L_3 <- 3
```

We keep the same explanatatory variable $X$. However, the $\beta$ matrix is now equal to : $$\left(\begin{array}{ccc} 5 & 2 & 10\\ 1/4 & -1 & 4\\ 6 & -3 & -5 \\ 1 & 3 & 5\end{array}\right)$$ 
  
and the $\Sigma$ matrix is $$\left(\begin{array}{ccc} 12 & 10 & 10\\ 10 & 15 & 10 \\ 10 & 10 & 20\end{array}\right)$$ 
  
  
```{r}
beta_true_3 <- matrix(c(5, 2, 10, 0, -1, 0, 6, -3, -5, 1, 0, 5), byrow = T,
                      nrow = p_simu, ncol = L_3)
Sigma_3 <- matrix(10, nrow = L_3, ncol = L_3)
diag(Sigma_3) <- c(12, 15, 20)
```

The $\Gamma$ matrix is $$\left(\begin{array}{ccc} 0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 0\end{array}\right)$$ 
  
```{r}
GAMMA_3 <- matrix(c(0, 0, 0, 0, 0, 0, 0, 0, 0),
                  nrow = L_3, ncol = L_3)
```

The $\rho$ matrix is $$\left(\begin{array}{ccc} 0.75 & 0.1 & 0.1 \\ 0.1 & 0.25 & 0.1 \\ 0.1 &0.1 & 0.8\end{array}\right)$$ 
  
```{r}
RHO_3 <- matrix(c(0.5, 0.1, 0.1, 0.1, 0.35, 0.1, 0.1, 0.1, 0.4),
                nrow = L_3, ncol = L_3)
```


**Simulation of a "N" distribution**
  
```{r}
y_N_3d <- simu_spatial_multi_y(X = x_simu, beta_true = beta_true_3, 
                               method_simulate = "N", Sigma = Sigma_3, 
                               GAMMA = GAMMA_3, RHO = RHO_3, W = W_simu) 
mapMAP@data[, c("y_1_3d", "y_2_3d", "y_3_3d")] <- y_N_3d
```

We plot the three component of $Y$ on the map: 
  
```{r, fig.width = 12, fig.height = 6}
op <- par(mfrow = c(1, 3), oma = c(0, 0, 0, 0), mar = c(0, 0, 1, 0))
choroLayer(spdf = mapMAP, var = "y_1_3d", legend.pos = "topleft", 
           method = "quantile")
choroLayer(spdf = mapMAP, var = "y_2_3d", legend.pos = "topleft", 
           method = "quantile")
choroLayer(spdf = mapMAP, var = "y_3_3d", legend.pos = "topleft", 
           method = "quantile")
par(op)
```


# Estimation #

## Estimation of the parameters by a 2SLS method

The function *estimate_spatial_multi_N()* estimates the coefficients associated to the multivariate Gaussian SAR model. The algorithm is based on Kelejian and Prucha (1998).

Input arguments are :
  
* **Y**, a matrix of size $n \times L$
  
* **X**, a matrix of explanatory variables of size $n \times K$,

* **W**, a spatial weight matrix of size $n \times n$,

* **GAMMA_esti**, a boolean which indicates if we estimate or nor the parameter associated to $\Gamma$.

The function returns a list with : 
  
* the estimate of the $\beta$ parameters

* the estimate of the $\Gamma$ matrix 

* the estimate of the $R$ matrix 

* the estimate of the $\Sigma$ matrix

To load the function:
```{r}
source("./R/estimate_spatial_multi_N.R")
source("./R/estimate_spatial_multi_gen_N.R")
```


### Examples: ###

**Simulated data when $L = 2$** 
```{r}
(res_multi_N <- estimate_spatial_multi_N(Y = y_N, X = x_simu, W = W_simu))
```

```{r}
(res_multi_N <- estimate_spatial_multi_N(Y = y_N, X = x_simu, W = W_simu, 
                                         GAMMA_esti = T))
```

**Simulated data when $L = 2$** 
  
```{r}
(res_multi_N_3D <- estimate_spatial_multi_N(Y = y_N_3d, X = x_simu, 
                                            W = W_simu))
```

```{r}
(res_multi_N_3D <- estimate_spatial_multi_N(Y = y_N_3d, X = x_simu, W = W_simu, 
                                            GAMMA_esti = T))
```


```{r}
(res_multi_N_gen_3D <- estimate_spatial_multi_gen_N(Y = y_N_3d, X = x_simu, 
  W = W_simu, ind_beta = matrix(c(T, F, T, T, T, T, T, F, T, F, T, T), 4, 3),
  ind_RHO = matrix(c(T, T, T, T, T, F, F, T, T), 3, 3),
  ind_GAMMA = matrix(c(F, T, T, T, F, F, F, T, F), 3, 3)))
```

## Application to the real data

We first load the data:
```{r}
source("R/preparation_base_ilr.R")
```

Then, we plot the data.
```{r}
Ye <- as(y_ilr, "matrix")
plot(Ye[,1], Ye[,2], xlab = expression(paste("ilr","_"[1],'(Z)')), 
     ylab = expression(paste("ilr","_"[2],'(Z)')), 
     xaxs = "i", yaxs = "i")
```

We prepare the explanatory variables:
  
```{r}
Xe <- as(cbind(1, x2_df[, c("age3_ilr1", "age3_ilr2", 
                            "unemp_rate", "income_rate")]),
         "matrix")
```

### Multivariate Gaussian model

We estimate first a multivariate gaussian model by using the *lm()* function.

```{r}
res_N <- lm(Ye ~ Xe - 1) 
```

Then, we look the spatial distribution of the residuals. For this, we first compute a spatial weight matrix based on the 4-nearest neighbours. 

```{r}
coords_fr <- coordinates(dep.2015.spdf)
W_listw <- nb2listw(knn2nb(knearneigh(coords_fr, 4)), 
                    style = "W")
W_dep <- listw2mat(W_listw)
```

We test the spatial autocorrelation in the residuals component by component:
```{r}
moran.mc(residuals(res_N)[, 1], listw = W_listw, nsim = 1000)
moran.mc(residuals(res_N)[, 2], listw = W_listw, nsim = 1000)
```

We plot the residuals:
```{r}
dep.2015.spdf@data[, c("res_y_1", "res_y_2")] <- residuals(res_N)
```


```{r, fig.width = 12, fig.height = 6}
library("cartography")
op <- par(mfrow = c(1, 2), oma = c(0, 0, 0, 0), mar = c(0, 0, 1, 0))
choroLayer(spdf = dep.2015.spdf, var = "res_y_1", legend.pos = "topleft", 
           method = "quantile", legend.values.rnd = 2)
choroLayer(spdf = dep.2015.spdf, var = "res_y_2", legend.pos = "topleft", 
           method = "quantile", legend.values.rnd = 2)
par(op)
```

We estimate a multivariate gaussian SAR model by using the *lm()* function.

```{r}
(res_sar_N <- estimate_spatial_multi_N(Ye, Xe, W_dep, GAMMA_esti = F))
```


# References 

*  Goulard M., Laurent T. and Thomas-Agnan C. (2017). About predictions in spatial autoregressive models: optimal and almost optimal strategies, Spatial Economic Analysis, 12:2-3, 304-325, DOI: 10.1080/17421772.2017.1300679 

* Nguyen T.H.A, Ruiz-Gazen, A., Thomas-Agnan C. and T. Laurent (2019). Multivariate Student versus Multivariate Gaussian Regression Models with Application to Finance. Journal of Risk and Financial Management, 12(1), 28.

